Naturalness and Long-lived Beyond the Standard Model Particles

### 2017-07-21 tex/draft.tex
Little TeX fixes - version sent to everybody

\usepackage[draft]{fixme}
\fxsetup{layout=footnote, marginclue}
\usepackage{amsmath,amsthm,amssymb,mathtools,graphicx,caption,subcaption,mathrsfs,cancel,multirow,enumitem,setspace,sectsty,cite,hyperref}
\usepackage{fullpage}

@@

\newpage
\tableofcontents
\newpage
\doublespacing

@@

    Experimentally, the mass of the Higgs boson is 125 GeV, a value that must be achieved by a linear combination of the bare mass and loop diagram corrections. The scale of the loop corrections is proportional to the square of the cutoff scale of the SM, which is the highest energy at which the SM is valid. If, hypothetically, the SM is valid all the way up to the Planck scale (where quantum gravity becomes important), then Higgs mass corrections would be order($MP^2$) ~ $10^{32}$ GeV.

    Order($10^{32}$) corrections imply that the bare Higgs mass must be negative and also order($10^{32}$) to reproduce the experimetnally known value. In fact, the bare mass and corrections must match to 1 part in ~$10^{30}$. While it's entirely possible that these two numbers just so happen to match so perfectly, but many \fxnote{weak} feel that such miraculous fine tuning must be a hint that some deeper physical mechanism is at work. That is, either some previously unknown symmetry is protecting the Higgs mass or the cutoff scale is order(TeV).

@@

    \fxnote{Optional paragraph about non-Higgs unnaturalness:    There are other unnatural quantities that people worry about in physics. I know of two off the top of my head: (1) thetaQCD in the strong CP problem (2) the cosmological consant. The strong CP problem comes from the fact that QCD technically allows CP violation even though such violation has never been observed. The degree of violation is parameterized as thetaQCD, which recasts the problem into the naturalness context of 'why is thetaQCD so small?'. The cosmological constant is experimentally ~60 to ~120 orders of magnitude smaller than the predicted value \fxnote{predicted how?}. While these issues deserve serious attention, they will not be considered further here}

@@

\section{Long-lived Particles}
    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and \fxnote{and/or?} produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict long-lived particles (LLPs). That is, particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point. This category includes everything from particles that decay at \fxnote{give ballparky number (~100um?)} to particles that propagate through the entire detector.

@@

\subsection{Tracker}
    \fxnote{cite CMS TDR, jinst 12 c02033}
    \fxnote{maybe I should trade some number listing for talking about how silicon detectors work}
    \fxnote{what about readout, cooling, etc?}
    In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions: at r < 20 cm, the high particle flux \fxnote{specify} requires small pixels to keep occupancy near 0.1\%. Moving farther from the IP decreases particle flux, which allows the use of larger silicon strips.

    The pixel tracker, which was upgraded in winter 2016/17, makes up the innermost layer and features approximately 124 million silicon pixels, each of which covers an active area of 100 by 150 $um^2$. The pixels are spread between four barrel layers (BPIX) at 2.9 < r < 16 cm and 3 endcap rings (FPIX) at 29 < z < 52 cm. The upgraded pixel tracker should provide 4-hit tracking out to eta < 2.4 and improved b-tagging over the original pixel tracker despite the increased pileup expected in 2017.

    Just outside the pixel tracker, the silicon strip tracker (SST) occupies the 20 < r < 116 cm and |z| < 282 cm region of CMS. Composed of 9.3 million silicon strips that vary from \_ to \_ in size \fxnote{look up} depending on particle flux in the region, the SST covers an active area of roughly 196 $m^2$. At each trigger, the strip output voltage is transferred via optical link to be read out by off-detector front-end electronics \fxnote{right jargon?}.

@@

    The crystals in each endcap section (EE) are arranged in an x-y grid that starts at |z| = 315 cm and covers a pseudorapidity range of 1.479 to 3.0. Additionally, a preshower detector, which consists of two planes of silicon strip detectors and a lead absorber, sits just inside the grid of lead tungstate crystals. \fxnote{Say what the preshower is for? Is it for pi0 rejection? If so, how and why?}

@@

\subsection{Muon System}
    \fxnote{cite CMS TDR}
    \fxnote{Specifically mention CMS design specs to meet physics goals?}
    The muon system (MS), which is the outermost subdetector, detects muons with 3 varieties of gaseous ionization detectors: drift tubes (DTs) in the barrel, cathode strip chambers (CSCs) in the endcaps, and resistive plate chambers (RPCs) in both the barrel and endcaps. The choice of detector varies between barrel and endcap due to the difference in neutron-induced background, muon rates, and magnetic field strength. The RPCs are useful in both the endcaps and barrel because their fast response and good time resolution allow unambiguous bunch crossing identification. Combining MS and tracker measurements improves the muon pT resolution at high momenta where the tracker pT resolution declines \fxnote{quantify or show std plot?}.

@@

\subsection{Trigger}
    \fxnote{cite CMS TDR, jinst 12 c03021}
    \fxnote{Is entire detector read out at L1?}
    \fxnote{Mention fpgas in L1 or more details about HLT?}
    \fxnote{Something about particle flow?}
    \fxnote{Sloppy, but I'm getting tired}
    The trigger reduces the data writing rate from the 40 MHz collision rate to less than 1 kHz so that events can be written to tape. The rate reduction happens in two stages: Level-1 (L1) and High-Level Trigger (HLT). L1 uses input from ECAL, HCAL, and MS that represents physics objects (e/gamma, tau, jets, MET, and mu) and uses custom electronics to reduce the rate to ~100 kHz in 3.8 us. The HLT then uses a dedicated processor farm to further reduce the rate to the desired < 1 kHz.

@@

    Slow and stopped particle searches are both sensitive to LLPs that are detector-stable but far more massive than any stable SM particle. Searches for low-velocity particles use ionization energy loss per unit distance (dE/dX) and time-of-flight (TOF) measurements to look for particles that are moving more slowly than the expected SM particles \fxnote{redundant}, which normally \fxnote{quantify} have beta > 0.9. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomenon that could be possible with long lifetimes and small initial kinetic energy. These analyses look for calorimeter energy deposits or hits in the muon system that are out of time with proton bunch crossings that could be due to the decay of stopped particles.

@@

    Disappearing tracks canditates, called pixel tracklets, are selected with a looser track reconstruction algorithm that runs over hits that were not associated to tracks by the standard reconstruction algorithm. The pixel tracklets are required to pass several isolation, pT, and quality requirements such as number of pixel layer hits. Finally, the disappearing criteria is enforced by requiring exactly zero associated hits in the subsequent tracking layer. Event selection requires one pixel tracklet, no electron or muon candidates (to suppress ttbar W/Z+j processes), and the MET and jet requirements associated with either of the two production modes described above.

@@

\subsection{Stopped Particles}
    \fxnote{cite CMS PAS EXO-17-004}
    \fxnote{mention how stopped particles complement HSCP search?}
    \fxnote{too redundant with overview?}
    With long enough lifetimes and sufficient mass, LLPs could come to rest in the detector before decaying. The eventual decay would then lead to energy deposits in the calorimeters or hits in the muon system that could be out of time with proton bunch crossings. A recent CMS analysis looked for such a signal in dimuon final states \fxnote{necessarily dimuon?} using the full 2015 and 2016 datasets.

@@

\subsection{Heavy Stable Charged Particles}
    \fxnote{cite CMS EXO7-010}
    \fxnote{nice dE/dx plot (fig 1})
    Particles that live for more than a couple ns \fxnote{Check this and be more exact} will appear stable in CMS and ATLAS because they can traverse the entire detector before decaying. If such a particle is also sufficiently massive, it can be catigorized as a heavy stable charged particle (HSCP). HSCPs are characterterized by a greater rate of energy loss via ionization due to the lower speed with which they pass through detector material. Additionally, HSCPs can be fractionally, singly, or multiply charged, which also affects their rate of energy loss.

    HSCPs will often be misidentified or unobserved by standard reconstruction algorithms and analysis criteria, which tend to assume beta~1 and |q|=+-e \fxnote{dE/dx plot is nice around here}. Inverting these assumptions provides two handles for a dedicated HSCP search in the context of three benchmark models: (1) HSCPs that form R-hadrons that either change the sign of their electric charge or become neutral before reaching the MS, (2) lepton-like HSCPs (quasi-stable sleptons) that are produced either through decays of squarks or gluinos or directly if squarks and gluinos are kinematically inaccessible, and (3) long-lived leptonlike fermions with aritrary electric charge. This discussion will focus on the first two models, as they bear more directly on naturalness and SUSY.

    \fxnote{I could talk about how low-beta tracks will show up as zig-zags in DTs b/c reconstruction assumes beta~1. Zig-zag offset is used to estimate beta. Not sure where to fit it in.} \fxnote{Worth mentioning tracker-only vs tracker+TOF}

>>>>

\usepackage[draft]{fixme}
\fxsetup{layout=footnote}
\usepackage[left]{lineno}
\usepackage{amsmath,amsthm,amssymb,mathtools,graphicx,caption,subcaption,mathrsfs,cancel,multirow,enumitem,setspace,sectsty,cite,hyperref}
\usepackage{fullpage}

@@

\newpage
\tableofcontents
\newpage
\doublespacing
\linenumbers

@@

    Experimentally, the mass of the Higgs boson is 125 GeV, a value that must be achieved by a linear combination of the bare mass and loop diagram corrections. The scale of the loop corrections is proportional to the square of the cutoff scale of the SM, which is the highest energy at which the SM is valid. If, hypothetically, the SM is valid all the way up to the Planck scale (where quantum gravity becomes important), then Higgs mass corrections would be order($MP^2$) $\approx 10^{32}$ GeV.

    Order($10^{32}$) corrections imply that the bare Higgs mass must be negative and also order($10^{32}$) to reproduce the experimetnally known value. In fact, the bare mass and corrections must match to 1 part in $\approx 10^{30}$. While it's entirely possible that these two numbers just so happen to match so perfectly, many \fxnote{weak} feel that such miraculous fine tuning must be a hint that some deeper physical mechanism is at work. That is, either some previously unknown symmetry is protecting the Higgs mass or the cutoff scale is order(TeV).

@@

    \fxnote{Optional paragraph about non-Higgs unnaturalness:    There are other unnatural quantities that people worry about in physics. I know of two off the top of my head: (1) thetaQCD in the strong CP problem (2) the cosmological consant. The strong CP problem comes from the fact that QCD technically allows CP violation even though such violation has never been observed. The degree of violation is parameterized as thetaQCD, which recasts the problem into the naturalness context of 'why is thetaQCD so small?'. The cosmological constant is experimentally $\approx$60 to $\approx$120 orders of magnitude smaller than the predicted value \fxnote{predicted how?}. While these issues deserve serious attention, they will not be considered further here}

@@

\section{Long-lived Particles}
    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and \fxnote{and/or?} produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict long-lived particles (LLPs). That is, particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point. This category includes everything from particles that decay at \fxnote{give ballparky number ($\approx$100um?)} to particles that propagate through the entire detector.

@@

\subsection{Tracker}
    \fxnote{cite CMS TDR, jinst 12 c02033}
    \fxnote{maybe I should trade some number listing for talking about how silicon detectors work}
    \fxnote{what about readout, cooling, etc?}
    In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions: at r $<$ 20 cm, the high particle flux \fxnote{specify} requires small pixels to keep occupancy near 0.1\%. Moving farther from the IP decreases particle flux, which allows the use of larger silicon strips.

    The pixel tracker, which was upgraded in winter 2016/17, makes up the innermost layer and features approximately 124 million silicon pixels, each of which covers an active area of 100 by 150 $um^2$. The pixels are spread between four barrel layers (BPIX) at 2.9 $<$ r $<$ 16 cm and 3 endcap rings (FPIX) at 29 $<$ z $<$ 52 cm. The upgraded pixel tracker should provide 4-hit tracking out to eta $<$ 2.4 and improved b-tagging over the original pixel tracker despite the increased pileup expected in 2017.

    Just outside the pixel tracker, the silicon strip tracker (SST) occupies the 20 $<$ r $<$ 116 cm and $|z|$ $<$ 282 cm region of CMS. Composed of 9.3 million silicon strips that vary from \_ to \_ in size \fxnote{look up} depending on particle flux in the region, the SST covers an active area of roughly 196 $m^2$. At each trigger, the strip output voltage is transferred via optical link to be read out by off-detector front-end electronics \fxnote{right jargon?}.

@@

    The crystals in each endcap section (EE) are arranged in an x-y grid that starts at $|z|$ = 315 cm and covers a pseudorapidity range of 1.479 to 3.0. Additionally, a preshower detector, which consists of two planes of silicon strip detectors and a lead absorber, sits just inside the grid of lead tungstate crystals. \fxnote{Say what the preshower is for? Is it for pi0 rejection? If so, how and why?}

@@

\subsection{Muon System}
    \fxnote{cite CMS TDR}
    \fxnote{Specifically mention CMS design specs to meet physics goals?}
    The muon system (MS), which is the outermost subdetector, detects muons with 3 varieties of gaseous ionization detectors: drift tubes (DTs) in the barrel, cathode strip chambers (CSCs) in the endcaps, and resistive plate chambers (RPCs) in both the barrel and endcaps. The choice of detector varies between barrel and endcap due to the difference in neutron-induced background, muon rates, and magnetic field strength. The RPCs are useful in the endcaps and barrel because their fast response and good time resolution allow unambiguous bunch crossing identification. Combining MS and tracker measurements improves the muon pT resolution at high momenta where the tracker pT resolution declines \fxnote{quantify or show std plot?}.

@@

\subsection{Trigger}
    \fxnote{cite CMS TDR, jinst 12 c03021}
    \fxnote{Is entire detector read out at L1?}
    \fxnote{Mention fpgas in L1 or more details about HLT?}
    \fxnote{Something about particle flow?}
    \fxnote{Sloppy, but I'm getting tired}
    The trigger reduces the data writing rate from the 40 MHz collision rate to less than 1 kHz so that events can be written to tape. The rate reduction happens in two stages: Level-1 (L1) and High-Level Trigger (HLT). L1 uses input from ECAL, HCAL, and MS that represents physics objects (e/gamma, tau, jets, MET, and mu) and uses custom electronics to reduce the rate to $\approx$100 kHz in 3.8 us. The HLT then uses a dedicated processor farm to further reduce the rate to the desired $<$ 1 kHz.

@@

    Slow and stopped particle searches are both sensitive to LLPs that are detector-stable but far more massive than any stable SM particle. Searches for low-velocity particles use ionization energy loss per unit distance (dE/dX) and time-of-flight (TOF) measurements to look for particles that are moving more slowly than the expected SM particles \fxnote{redundant}, which normally \fxnote{quantify} have beta $>$ 0.9. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomenon that could be possible with long lifetimes and small initial kinetic energy. These analyses look for calorimeter energy deposits or hits in the muon system that are out of time with proton bunch crossings that could be due to the decay of stopped particles.

@@

    Disappearing track candidates, called pixel tracklets, are selected with a looser track reconstruction algorithm that runs over hits that were not associated to tracks by the standard reconstruction algorithm. The pixel tracklets are required to pass several isolation, pT, and quality requirements such as number of pixel layer hits. Finally, the disappearing criteria is enforced by requiring exactly zero associated hits in the subsequent tracking layer. Event selection requires one pixel tracklet, no electron or muon candidates (to suppress ttbar W/Z+j processes), and the MET and jet requirements associated with either of the two production modes described above.

@@

\subsection{Stopped Particles}
    \fxnote{cite CMS PAS EXO-17-004}
    \fxnote{mention how stopped particles complement HSCP search?}
    \fxnote{too redundant with overview?}
    With long enough lifetimes and sufficient mass, LLPs could come to rest in the detector before decaying. The eventual decay would then lead to energy deposits in the calorimeters or hits in the muon system that could be out of time with proton bunch crossings. A recent CMS analysis looked for such a signal in dimuon final states using the full 2015 and 2016 datasets.

@@

\subsection{Heavy Stable Charged Particles}
    \fxnote{cite CMS EXO7-010}
    \fxnote{nice dE/dx plot (fig 1)}
    Particles that live for more than a couple ns \fxnote{Check this and be more exact} will appear stable in CMS and ATLAS because they can traverse the entire detector before decaying. If such a particle is also sufficiently massive, it can be catigorized as a heavy stable charged particle (HSCP). HSCPs are characterterized by a greater rate of energy loss via ionization due to the lower speed with which they pass through detector material. Additionally, HSCPs can be fractionally, singly, or multiply charged, which also affects their rate of energy loss.

    HSCPs will often be misidentified or unobserved by standard reconstruction algorithms and analysis criteria, which tend to assume beta$\approx$1 and $|q|$ = e. Inverting these assumptions provides two handles for a dedicated HSCP search in the context of three benchmark models: (1) HSCPs that form R-hadrons that either change the sign of their electric charge or become neutral before reaching the MS, (2) lepton-like HSCPs (quasi-stable sleptons) that are produced either through decays of squarks or gluinos or directly if squarks and gluinos are kinematically inaccessible, and (3) long-lived leptonlike fermions with arbitrary electric charge. This discussion will focus on the first two models, as they bear more directly on naturalness and SUSY.

    \fxnote{I could talk about how low-beta tracks will show up as zig-zags in DTs b/c reconstruction assumes beta$\approx$1. Zig-zag offset is used to estimate beta. Not sure where to fit it in.} \fxnote{Worth mentioning tracker-only vs tracker+TOF}
